\huge \bf {Using Optimal Time Step Selection to \\ Boost the Accuracy of FD Schemes for \\ Variable Coefficient PDEs }

### 2009-03-20 papers/OTSPDE-WCE2009/OTSPDE-WCE2009.tex
OTSPDE: - Checking in Jim's revisions to WCE2009 manuscript.

% >>>>>>>>>>>>>>>>>>>>>>> Author's Name, Thanks or Affliation <<<<<<<<
\author{Kevin T. Chu\thanks{
  K.T. Chu gratefully acknowledges the support of Vitamin D, Inc.
  and the Institute for High-Performance Computing (IHPC) in Singapore. 
  Vitamin D, Inc., Menlo Park, CA 94025 \& 
  Institute of High Performance Computing, A*STAR, Singapore, Singapore.
  Email: ktchu@serendipityresearch.org.
  Manuscript submitted 2008 March 2. 
 } 
 \ and
 James V. Lambers\thanks{Department of Energy Resources Engineering, 
 Stanford University, Stanford, CA 94305.  Email: lambers@stanford.edu.}
}

@@

% >>>>>>>>>>>>>>>>>>>>>>>>> Keywords and Abstract <<<<<<<<<<<<<<<<<<<<<
{\hspace{1pc} {\it{\small Abstract}}{\bf{\small---We extend the technique 
of optimal time step (OTS) selection for finite difference (FD) schemes of time 
dependent PDEs to PDEs where the leading-order spatial derivative term has a
spatially varying coefficient.  The basic approach involves identifying a 
transformation of the domain that eliminates the spatial dependence of the 
coefficient for the leading-order term.  This change of variables is then used 
to define an optimal computational grid for the FD scheme on the 
\emph{original} domain.   By using both the optimal grid and OTS, we are
able to boost the order of accuracy above what would be expected from a formal
analysis of the FD scheme.  We illustrate the utility of our method by 
applying it to variable coefficient wave and diffusion equations.  In 
addition, we demonstrate the viability of OTS selection to the two-step 
Kreiss-Petersson-Ystr\"om discretization of the wave equation.

@@

% >>>>>>>>>>>>>>>>>>>>>> START OF PAPER <<<<<<<<<<<<<<<<<<<<<<<<<<<<<<
\section{Introduction}
\label{Introduction}
Optimal time step (OTS) selection is a surprisingly simple and effective way to 
boost the order of accuracy of finite difference (FD) schemes for 
time-dependent partial differential equations~\cite{chu_otspde}.  The basic
principle underlying OTS selection is that a careful choice of time step (and 
the addition of a few correction terms) can boost the \emph{order} of 
accuracy for formally low-order finite difference schemes.  For instance, it 
is possible to obtain a fourth-order accurate solution of the 1D diffusion 
equation using only the standard second-order central difference approximation 
and forward Euler time integration.  As demonstrated in~\cite{chu_otspde}, OTS 
selection works for linear and semilinear PDEs in any number of space 
dimensions on both regular and irregular domains.

Unfortunately, a limitation of the existing formulation of OTS selection is 
the requirement that the leading-order spatial derivative of the PDE be 
constant coefficient.  In this paper, we remove this limitation and extend OTS 
selection to be applicable to general variable coefficient semilinear PDEs.  
Our approach is to identify a transformation of the domain which 
eliminates the spatial dependence of the coefficient on the leading-order 
spatial derivative.  Using this transformation, we define an optimal 
computational grid which can be used in conjunction with OTS selection to 
boost the order of accuracy of formally low-order FD schemes for the PDE.  
Optimal grid selection is an important extension of the philosophy that the 
accuracy of numerical methods can be boosted via optimization of the 
parameters used to compute the solution.

This paper is organized as follows.  First, we briefly review OTS selection 
for PDEs with a constant coefficient leading-order spatial derivative term.  
Next, we show how to combine optimal grid and optimal time step selection to 
boost the accuracy of finite difference schemes for variable coefficient PDEs.  
We then demonstrate the use of variable coefficient OTS selection on the 
variable coefficient wave and heat equations.  Our analysis of the wave is 
equation is of particular note because it is \emph{not} based on a single step 
time integration scheme.  We conclude with a summary of our main results
and thoughts on future directions for research.

\section{Review of OTS Selection for Constant Coefficient PDEs
  \label{sec:ots_const_coef_pdes}}
Optimal time step selection is not by itself a method for constructing 
finite difference schemes.  Rather, it is a technique for enhancing the 
performance of existing FD schemes by carefully choosing the time step to 
eliminate low-order numerical errors.  There are two fundamental ideas 
underlying OTS selection.  First, a judicious choice of time step can be 
used to eliminate the leading-order terms in the error.  Second, the PDE 
provides valuable insight into the discretization errors for FD schemes.  
Combining these two simple ideas often yields an optimal time step which 
can be used to \emph{boost} the order of accuracy of a given FD scheme 
above what would be expected from a formal analysis of the scheme.

To illustrate the basic theory behind OTS selection, let us consider the 
constant coefficient diffusion equation:
\beq
  u_{t} = D u_{xx} + f(x,t),
 \label{eq:diffusion_eqn_1d}
\eeq
where $D$ is the diffusion constant and $f(x,t)$ is a source term.  Perhaps
the simplest FD scheme for this equation uses forward Euler time integration 
and the standard second-order central difference approximation for the 
Laplacian:
\beq
  u^{n+1}_j = u^{n}_j
    + \dt
      \left( D \left[\frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2}\right]
           + f^n_j \right).
 \label{eq:diffusion_eqn_1d_FD_scheme}
\eeq
This scheme is formally first-order in time and second-order in 
space.  The stability constraint $\dt \le \dx^2/2D$ implies that the scheme is
$O(\dx^2)$ accurate overall.

OTS selection is based on a detailed analysis of the leading-order errors in 
FD schemes.  So, we begin by deriving the truncation equation for the scheme 
(\ref{eq:diffusion_eqn_1d_FD_scheme}).  Employing Taylor series expansions and 
the PDE (\ref{eq:diffusion_eqn_1d}), we find that the true solution satisfies
\bea
  \tu^{n+1}_j &=& \tu^{n}_j
  + \dt \left( D \tu_{xx} + f \right)
  \nonumber \\
  &+& \frac{\dt^2}{2} \left( D^2 \tu_{xxxx} + D f_{xx} + f_t \right)
  + O \left( \dt^3 \right) \ \ 
  \label{eq:diffusion_eqn_1d_time_err}
\eea
and that the central difference approximation for the Laplacian satisfies
\beq
  \frac{\tu^{n}_{j+1} -2 \tu^{n}_j + \tu^{n}_{j-1}}{\dx^2}  =
  \tu_{xx} + \frac{\dx^2}{12} \tu_{xxxx}
  + O(\dx^4)
  \label{eq:diffusion_eqn_1d_space_err}
\eeq
Therefore, the truncation error for (\ref{eq:diffusion_eqn_1d_FD_scheme})
is given by
\bea
  & & \tu_{xxxx}
  \left[ \frac{\dx^2}{12} - \frac{D \dt}{2} \right] (D \dt)
  - \frac{\dt^2}{2} \left( D f_{xx}
  + f_t \right)
  \nonumber \\
  & & + \ O(\dt \dx^4) + O(\dt^3).
  \label{eq:diffusion_eqn_1d_trunc_err}
\eea
From this expression, we see that choosing the time step to be
$\dt = \dx^2/6D$ and adding the correction term 
\beq
  \frac{\dt^2}{2} \left( D f_{xx} + f_t \right)
\eeq
eliminates the leading-order truncation error.  Using the heuristic
argument that the global error is equal to the local error divided by 
$\dt$~~\cite{gko_book}, we find that the numerical solution is 
$O(\dx^4) + O(\dt^2) = O(\dx^4)$ accurate -- the FD scheme has been
boosted from second- to fourth-order accuracy.

@@

\section{OTS Selection for Variable Coefficient PDEs
  \label{sec:ots_var_coef_pdes}}
Transformation of the spatial domain is a natural way to extend OTS selection 
to semilinear PDEs that have variable coefficient leading-order spatial 
derivatives.  Consider a semilinear PDE of the form:
\bea
  \frac{\partial u}{\partial t} = a^n(x) \frac{\partial^n u}{\partial x^n}
 + F \left( \frac{\partial^{n-1} u}{\partial x^{n-1}}, \ldots,
           \frac{\partial u}{\partial x}, u \right)
 + f(x,t)
\label{eq:semilinear_PDE}
\eea
on the domain $0 < x < 1$.  Using the change of variables\footnote{This
transformation is a generalization of the change of variables proposed
in~\cite{guidotti2006} for the wave equation.}
\beq
  y = \Phi(x) = \abar \int_0^x \frac{1}{a(\xi)} d\xi
  \label{eq:change_of_variables}
\eeq
with 
\beq
  \abar \equiv \left( \int_0^1 \frac{1}{a(\xi)} d\xi \right)^{-1},
  \label{eq:abar}
\eeq
we can completely eliminate the spatial dependence of the coefficient on the 
leading-order spatial derivative (at the expense of adding 
variable coefficient lower-order spatial derivative terms).  The PDE 
(\ref{eq:semilinear_PDE}) on the transformed domain has the form
\bea
  \frac{\partial u}{\partial t} = {\abar}^n \frac{\partial^n u}{\partial y^n}
 + \tilde{F} \left( \frac{\partial^{n-1} u}{\partial y^{n-1}},
           \ldots,
           \frac{\partial u}{\partial y}, u \right)
 + f(y,t),
\label{eq:semilinear_PDE_transformed}
\eea
where $\tilde{F}$ includes any lower-order terms introduced by the process of 
changing variables.

@@

\subsection{Derivation of OTS and Correction Terms for Optimal Grid
  \label{sec:derivation_ots_corr_terms} }
To derive the optimal time step, we apply the procedure presented 
in~\cite{chu_otspde} to the PDE in the transformed domain 
(\ref{eq:semilinear_PDE_transformed}).  However, rather than analyzing the 
full PDE, we focus only on the simplified PDE containing only the time 
derivative term and leading-order spatial derivative:
\beq
  \frac{\partial u}{\partial t} = {\abar}^n \frac{\partial^n u}{\partial y^n}
  \label{eq:simplified_PDE_transformed}.
\eeq
When a uniform grid and standard finite difference schemes are used to solve
this PDE on the transformed domain, we find that the optimal time step is given
by $\dt_{opt} = \alpha (\dy / \abar)^n$, where $\alpha$ depends on the 
particular FD scheme used approximate (\ref{eq:simplified_PDE_transformed}).
Because the FD schemes on the original and transformed domains are chosen 
to be compatible, $\dt_{opt}$ is also the optimal time step to use when 
solving the PDE on the optimal grid.

@@

To boost the accuracy, we also need to derive the appropriate correction terms.
When working on the transformed domain, we simply follow the procedure 
described in~\cite{chu_otspde}.  However, when working on the original domain 
using the optimal mesh, the analysis is more complicated and generally depends
on the details of the FD scheme and PDE.  The main challenge is deciding which 
terms of the leading-order temporal error are not canceled out by the choice 
of time step.

Fortunately, a simple heuristic that can be used to derive the correction 
terms.  When deriving $u_{tt}$, each spatial derivative term in the PDE (on 
the original domain) gives rise to several potential correction terms.  We 
conjecture that of these candidates, we need only retain those terms involving 
spatial derivatives of $u$ with order not exceeding the order the originating 
spatial derivative term in the PDE.  As a concrete example, consider a term in 
the PDE involving $u_x$.  The corresponding correction terms should only 
involve $u_x$ and $u$ but not $u_{xx}$ or higher derivatives.  While this 
assertion has yet to be proven, it seems to hold for the PDEs we have examined.

@@

\subsection{Variable Coefficient Wave Equation}
In this section, we apply OTS and optimal grid selection to the
variable coefficient wave equation.  Because we choose a FD scheme based on 
the Kreiss-Petersson-Ystr\"om discretization scheme for the wave 
equation~\cite{kreiss2002}, we begin by analyzing OTS selection for the of 
the constant coefficient wave equation and demonstrate its applicability for 
this multistep method.

\subsubsection*{OTS Selection for the KPY Discretization of the 
  Constant-Coefficient Wave Equation}
In~\cite{kreiss2002}, Kreiss, Petersson, and Ystr\"om analyzed the following
direct discretization of the second-order wave equation, 
$u_{tt} - c^2 u_{xx} = f$,
without conversion to a first-order system of equations:
\bea
    \frac{u^{n+1}_i - 2 u^n_i + u^{n-1}_i}{\dt^2}
  = c^2 \left( \frac{u^{n}_{i+1} - 2 u^n_i + u^n_{i-1}}{\dx^2} \right)
  + f
  \label{eq:wave_eqn_KPY}
\eea
and showed that it is second-order accurate in space and time with a 
stability constraint of the form $\dt = O(\dx)$.

@@

To compute the optimal time step and correction terms for this scheme, we 
follow the procedure outlined in~\cite{chu_otspde} and 
Section~\ref{sec:ots_const_coef_pdes}.  We begin by deriving the truncation 
error for the scheme.  Employing Taylor series expansions, it is 
straightforward to show that the true solution satisfies:
\bea
  & &\frac{\tu^{n+1}_i - 2 \tu^n_i + \tu^{n-1}_i}{\dt^2}
    - \frac{\dt^2}{12} \tu_{tttt} + O(\dt^4)
  = \nonumber \\
  & & c^2 \left( \frac{\tu^{n}_{i+1} - 2 \tu^n_i + \tu^n_{i-1}}{\dx^2} 
  -\frac{\dx^2}{12} \tu_{xxxx} + O(\dx^4) \right)
  \nonumber \\
  & & +f
\eea
Combining this result with the observation that 
\beq
  \tu_{tttt} = c^4 \tu_{xxxx} + c^2 f_{xx} + f_{tt}
\eeq
we find that the truncation error is given by
\bea
    & & c^2 \frac{\dx^2}{12} \tu_{xxxx} 
    - \frac{\dt^2}{12} \left(c^4\tu_{xxxx} + c^2 f_{xx} + f_{tt} \right) 
    \nonumber \\
    & & + \ O(\dx^4) + O(\dt^4) 
\eea
Thus, we can eliminate the leading-order term in the discretization
error by choosing $\dt = \dx/c$ and adding the correction term 
\beq
  \frac{\dt^2}{12} \left(c^2 f_{xx} + f_{tt} \right) 
\eeq
to the right-hand side of (\ref{eq:wave_eqn_KPY}).

@@

\paragraph{High Accuracy Required for First Time Step}
For the KPY discretization of the wave equation, it is critical that a 
high-order accurate method is used to compute the first time step because 
the error introduced during this step affects the global error at all times.
Specifically, for the error of the KPY scheme to be $O(\dx^p)$, the numerical
solution at the first time step must be accurate to $O(\dx^{p+1})$.  In other 
words the error introduced by the first time step must be at least one order
of accuracy higher than desired for the overall solution.

We can understand this need for higher-order accuracy of the first time step 
by solving the difference equation for the normal modes of the error:
\bea
  \frac{\he^{n+1} - 2 \he^{n} + \he^{n-1}}
       {\dt^2} = \dt^2 \lambda \he^{n}
  \label{eq:error_eqn_normal_mode},
\eea
where $\he$ is the coefficient of an arbitrary normal mode of the spatial
operator for the error $e \equiv u - \tu$.  The solution of this equation 
is~\cite{carrier_pearson_book}
\bea
  \he^n = \he^1 \frac{\kappa_+^n - \kappa_-^n}{\kappa_+ - \kappa_-} 
      + \he^0 \kappa_+ \kappa_- 
        \frac{\kappa_+^{n-1} - \kappa_-^{n-1}}{\kappa_+ - \kappa_-} 
  \label{eq:error_eqn_normal_mode_soln}
\eea
where $\kappa_\pm$, the roots of the characteristic equation for 
(\ref{eq:error_eqn_normal_mode}), are given by~\cite{kreiss2002}
\beq
  \kappa_\pm = 1 + \frac{1}{2} \lambda \dt^2 
             \pm \dt \sqrt{\lambda + \frac{\lambda^2 \dt^2}{4}}.
\eeq
Notice that the denominator of both terms in 
(\ref{eq:error_eqn_normal_mode_soln}) is $O(\dt)$.  As a result, the 
global error is always at least one temporal order of accuracy less than the 
error in the initial conditions.  Therefore, the initial errors $e^0$ and 
$e^1$ must be $O(\dt^{p+1})$ in order to for the solution to be $O(\dt^p)$. 
For KPY without OTS selection, this observation indicates that the first
time step should be at least third-order accurate, which is exactly the way
that Kreiss \etal chose to compute the first time step in~\cite{kreiss2002}.
Because the OTS selection transforms KPY into a fourth-order accurate scheme, 
we require that the first time step is at least fifth-order accurate.

It is straightforward to construct a fifth-order approximation for the first 
time step by using a fifth-order Taylor series expansion in time:
\bea
  u^1 &=& u^0 + \dt u_t^0 + \frac{\dt^2}{2} u_{tt}^0
      + \frac{\dt^3}{6} u_{ttt}^0 
   + \frac{\dt^4}{24} u_{tttt}^0
  \nonumber \\
  &=& u^0 + \dt u_t^0 + \frac{\dt^2}{2} \left(c^2 u_{xx}^0 + f\right)
  \nonumber \\
  & & + \frac{\dt^3}{6} \left(c^2 \left(u_t\right)_{xx}^0 + f_t\right)
  \nonumber \\
  & & + \ \frac{\dt^4}{24} \left(c^4 u_{xxxx}^0 + c^2 f_{xx} + f_{tt}\right)
  \label{eq:KPY_fifth_order_first_step}
\eea
Note that sufficiently high-order accurate finite difference stencils may be 
used to compute some of the higher order terms in this expression.  However,
when advancing the solution using the optimal time step, special care must be 
exercised if finite differences are used to compute the derivatives in 
(\ref{eq:KPY_fifth_order_first_step}) because higher-order terms may be 
automatically (implicitly) included by lower-order terms in the expansion.

>>>>

% >>>>>>>>>>>>>>>>>>>>>>> Author's Name, Thanks or Affliation <<<<<<<<
\author{Kevin T. Chu\thanks{
  K.T. Chu gratefully acknowledges the support of Vitamin D, Inc.
  and the Institute for High-Performance Computing (IHPC) in Singapore. 
  Vitamin D, Inc., Menlo Park, CA 94025 \& 
  Institute of High Performance Computing, A*STAR, Singapore, Singapore.
  Email: ktchu@serendipityresearch.org.
  Manuscript submitted 2009 March 20. 
 } 
 \ and
 James V. Lambers\thanks{Department of Energy Resources Engineering, 
 Stanford University, Stanford, CA 94305.  Email: lambers@stanford.edu.}
}

@@

% >>>>>>>>>>>>>>>>>>>>>>>>> Keywords and Abstract <<<<<<<<<<<<<<<<<<<<<
{\hspace{1pc} {\it{\small Abstract}}{\bf{\small---We extend the technique 
of optimal time step (OTS) selection for finite difference (FD) schemes of time 
dependent PDEs to PDEs where the leading-order spatial derivative term has a
spatially varying coefficient.  The basic approach involves identifying a 
transformation of the domain that eliminates the spatial dependence of the 
coefficient for the leading-order term.  This change of variables is then used 
to define an optimal computational grid for the FD scheme on the 
\emph{original} domain.   By using both the optimal grid and OTS selection, we are
able to boost the order of accuracy above what would be expected from a formal
analysis of the FD scheme.  We illustrate the utility of our method by 
applying it to variable coefficient wave and diffusion equations.  In 
addition, we demonstrate the viability of OTS selection for the two-step 
Kreiss-Petersson-Ystr\"om discretization of the wave equation.

@@

% >>>>>>>>>>>>>>>>>>>>>> START OF PAPER <<<<<<<<<<<<<<<<<<<<<<<<<<<<<<
\section{Introduction}
\label{Introduction}
Optimal time step (OTS) selection is a surprisingly simple and effective way to 
boost the order of accuracy of finite-difference (FD) schemes for 
time-dependent partial differential equations~\cite{chu_otspde}.  The basic
principle underlying OTS selection is that a careful choice of time step (and 
the addition of a few correction terms) can boost the \emph{order} of 
accuracy for formally low-order finite difference schemes.  For instance, it 
is possible to obtain a fourth-order accurate solution of the 1D diffusion 
equation using only the standard second-order central difference approximation 
and forward Euler time integration.  As demonstrated in~\cite{chu_otspde}, OTS 
selection works for linear and semilinear PDEs in any number of space 
dimensions on both regular and irregular domains.

Unfortunately, a limitation of the existing formulation of OTS selection is 
the requirement that the leading-order spatial derivative of the PDE have a 
constant coefficient.  In this paper, we remove this limitation and extend OTS 
selection to general variable-coefficient semilinear PDEs.  
Our approach is to identify a transformation of the domain which 
eliminates the spatial dependence of the coefficient on the leading-order 
spatial derivative.  Using this transformation, we define an optimal 
computational grid which can be used in conjunction with OTS selection to 
boost the order of accuracy of formally low-order FD schemes for the PDE.  
Optimal grid selection is an important extension of the philosophy that the 
accuracy of numerical methods can be boosted via optimization of the 
parameters used to compute the solution.

This paper is organized as follows.  First, we briefly review OTS selection 
for PDEs with a constant-coefficient leading-order spatial derivative term.  
Next, we show how to combine optimal grid and optimal time step selection to 
boost the accuracy of finite difference schemes for variable-coefficient PDEs.  
We then demonstrate the use of variable-coefficient OTS selection on the 
variable-coefficient wave and heat equations.  Our analysis of the wave is 
equation is particularly noteworthy because it is \emph{not} based on a one-step 
time integration scheme.  We conclude with a summary of our main results
and thoughts on future directions for research.

\section{Review of OTS Selection for Constant Coefficient PDEs
  \label{sec:ots_const_coef_pdes}}
Optimal time step selection is not by itself a method for constructing 
finite-difference schemes.  Rather, it is a technique for enhancing the 
performance of existing FD schemes by carefully choosing the time step to 
eliminate low-order numerical errors.  There are two fundamental ideas 
underlying OTS selection.  First, a judicious choice of time step can be 
used to eliminate the leading-order terms in the error.  Second, the PDE 
provides valuable insight into the discretization errors for FD schemes.  
Combining these two simple ideas often yields an optimal time step which 
can be used to \emph{boost} the order of accuracy of a given FD scheme 
above what would be expected from a formal analysis of the scheme.

To illustrate the basic theory behind OTS selection, let us consider the 
constant-coefficient diffusion equation
\beq
  u_{t} = D u_{xx} + f(x,t),
 \label{eq:diffusion_eqn_1d}
\eeq
where $D$ is the diffusion constant and $f(x,t)$ is a source term.  Perhaps
the simplest FD scheme for this equation uses forward Euler time integration 
and the standard second-order central difference approximation for the 
Laplacian:
\beq
  u^{n+1}_j = u^{n}_j
    + \dt
      \left( D \left[\frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2}\right]
           + f^n_j \right).
 \label{eq:diffusion_eqn_1d_FD_scheme}
\eeq
This scheme is formally first-order in time and second-order in 
space.  The stability constraint $\dt \le \dx^2/2D$ implies that the scheme is
$O(\dx^2)$ accurate overall.

OTS selection is based on a detailed analysis of the leading-order errors in 
FD schemes.  We begin by deriving the truncation error for the scheme 
(\ref{eq:diffusion_eqn_1d_FD_scheme}).  Employing Taylor series expansions and 
the PDE (\ref{eq:diffusion_eqn_1d}), we find that the true solution satisfies
\bea
  \tu^{n+1}_j &=& \tu^{n}_j
  + \dt \left( D \tu_{xx} + f \right)
  \nonumber \\
  &+& \frac{\dt^2}{2} \left( D^2 \tu_{xxxx} + D f_{xx} + f_t \right)
  + O \left( \dt^3 \right) \ \ 
  \label{eq:diffusion_eqn_1d_time_err}
\eea
and that the central difference approximation for the Laplacian satisfies
\beq
  \frac{\tu^{n}_{j+1} -2 \tu^{n}_j + \tu^{n}_{j-1}}{\dx^2}  =
  \tu_{xx} + \frac{\dx^2}{12} \tu_{xxxx}
  + O(\dx^4).
  \label{eq:diffusion_eqn_1d_space_err}
\eeq
Therefore, the truncation error for (\ref{eq:diffusion_eqn_1d_FD_scheme})
is given by
\bea
  & & \tu_{xxxx}
  \left[ \frac{\dx^2}{12} - \frac{D \dt}{2} \right] (D \dt)
  - \frac{\dt^2}{2} \left( D f_{xx}
  + f_t \right)
  \nonumber \\
  & & + \ O(\dt \dx^4) + O(\dt^3).
  \label{eq:diffusion_eqn_1d_trunc_err}
\eea
From this expression, we see that choosing the time step to be
$\dt = \dx^2/6D$ and adding the correction term 
\beq
  \frac{\dt^2}{2} \left( D f_{xx} + f_t \right)
\eeq
eliminates the leading-order truncation error.  Using the heuristic
argument that the global error is equal to the local error divided by 
$\dt$~~\cite{gko_book}, we find that the numerical solution is 
$O(\dx^4) + O(\dt^2) = O(\dx^4)$ accurate -- the FD scheme has been
boosted from second- to fourth-order accuracy.

@@

\section{OTS Selection for Variable Coefficient PDEs
  \label{sec:ots_var_coef_pdes}}
Transformation of the spatial domain is a natural way to extend OTS selection 
to semilinear PDEs that have variable coefficient leading-order spatial 
derivatives.  Consider a semilinear PDE of the form:
\bea
  \frac{\partial u}{\partial t} = a^n(x) \frac{\partial^n u}{\partial x^n}
 + F \left( \frac{\partial^{n-1} u}{\partial x^{n-1}}, \ldots,
           \frac{\partial u}{\partial x}, u \right)
 + f(x,t)
\label{eq:semilinear_PDE}
\eea
on the domain $0 < x < 1$.  Using the change of variables\footnote{This
transformation is a generalization of the change of variables proposed
in~\cite{guidotti2006} for the wave equation.}
\beq
  y = \Phi(x) = \abar \int_0^x \frac{1}{a(\xi)} d\xi
  \label{eq:change_of_variables}
\eeq
with 
\beq
  \abar \equiv \left( \int_0^1 \frac{1}{a(\xi)} d\xi \right)^{-1},
  \label{eq:abar}
\eeq
we can completely eliminate the spatial dependence of the coefficient on the 
leading-order spatial derivative (at the expense of adding 
lower-order variable-coefficient spatial derivative terms).  The PDE 
(\ref{eq:semilinear_PDE}) on the transformed domain has the form
\bea
  \frac{\partial u}{\partial t} = {\abar}^n \frac{\partial^n u}{\partial y^n}
 + \tilde{F} \left( \frac{\partial^{n-1} u}{\partial y^{n-1}},
           \ldots,
           \frac{\partial u}{\partial y}, u \right)
 + f(y,t),
\label{eq:semilinear_PDE_transformed}
\eea
where $\tilde{F}$ includes any lower-order terms introduced by the process of 
changing variables.

@@

\subsection{Derivation of OTS Selection and Correction Terms for Optimal Grid
  \label{sec:derivation_ots_corr_terms} }
To derive the optimal time step, we apply the procedure presented 
in~\cite{chu_otspde} to the PDE in the transformed domain 
(\ref{eq:semilinear_PDE_transformed}).  However, rather than analyzing the 
full PDE, we focus only on the simplified PDE containing only the time 
derivative term and leading-order spatial derivative:
\beq
  \frac{\partial u}{\partial t} = {\abar}^n \frac{\partial^n u}{\partial y^n}
  \label{eq:simplified_PDE_transformed}.
\eeq
When a uniform grid and standard finite difference schemes are used to solve
this PDE on the transformed domain, we find that the optimal time step is given
by $\dt_{opt} = \alpha (\dy / \abar)^n$, where $\alpha$ depends on the 
particular FD scheme used approximate (\ref{eq:simplified_PDE_transformed}).
Because the FD schemes on the original and transformed domains are chosen 
to be compatible, $\dt_{opt}$ is also the optimal time step to use when 
solving the PDE on the optimal grid.

@@

To boost the accuracy, we also need to derive the appropriate correction terms.
When working on the transformed domain, we simply follow the procedure 
described in~\cite{chu_otspde}.  However, when working on the original domain 
using the optimal mesh, the analysis is more complicated and generally depends
on the details of the FD scheme and PDE.  The main challenge is deciding which 
terms of the leading-order temporal error are not cancelled out by the choice 
of time step.

Fortunately, a simple heuristic can be used to derive the correction 
terms.  When deriving $u_{tt}$, each spatial derivative term in the PDE (on 
the original domain) gives rise to several potential correction terms.  We 
conjecture that of these candidates, we need only retain those terms involving 
spatial derivatives of $u$ with order not exceeding the order the originating 
spatial derivative term in the PDE.  As a concrete example, consider a term in 
the PDE involving $u_x$.  The corresponding correction terms should only 
involve $u_x$ and $u$ but not $u_{xx}$ or higher derivatives.  While this 
assertion has yet to be proven, it seems to hold for the PDEs we have examined.

@@

\subsection{Variable Coefficient Wave Equation}
In this section, we apply OTS and optimal grid selection to the
variable coefficient wave equation.  Because we choose a FD scheme based on 
the Kreiss-Petersson-Ystr\"om discretization scheme for the wave 
equation~\cite{kreiss2002}, we begin by analyzing OTS selection for 
the constant-coefficient wave equation and demonstrate its applicability to
this multistep method.

\subsubsection*{OTS Selection for the KPY Discretization of the 
  Constant-Coefficient Wave Equation}
In~\cite{kreiss2002}, Kreiss, Petersson, and Ystr\"om analyzed the following
direct discretization of the second-order wave equation, 
$u_{tt} - c^2 u_{xx} = f$,
without conversion to a first-order system of equations:
\bea
    \frac{u^{n+1}_i - 2 u^n_i + u^{n-1}_i}{\dt^2}
  = c^2 \left( \frac{u^{n}_{i+1} - 2 u^n_i + u^n_{i-1}}{\dx^2} \right)
  + f.
  \label{eq:wave_eqn_KPY}
\eea
They showed that it is second-order accurate in space and time with a 
stability constraint of the form $\dt = O(\dx)$.

@@

To compute the optimal time step and correction terms for this scheme, we 
follow the procedure outlined in~\cite{chu_otspde} and 
Section~\ref{sec:ots_const_coef_pdes}.  We begin by deriving the truncation 
error for the scheme.  Employing Taylor series expansions, it is 
straightforward to show that the true solution satisfies
\bea
  & &\frac{\tu^{n+1}_i - 2 \tu^n_i + \tu^{n-1}_i}{\dt^2}
    - \frac{\dt^2}{12} \tu_{tttt} + O(\dt^4)
  = \nonumber \\
  & & c^2 \left( \frac{\tu^{n}_{i+1} - 2 \tu^n_i + \tu^n_{i-1}}{\dx^2} 
  -\frac{\dx^2}{12} \tu_{xxxx} + O(\dx^4) \right)
  \nonumber \\
  & & +f.
\eea
Combining this result with the observation that 
\beq
  \tu_{tttt} = c^4 \tu_{xxxx} + c^2 f_{xx} + f_{tt},
\eeq
we find that the truncation error is given by
\bea
    & & c^2 \frac{\dx^2}{12} \tu_{xxxx} 
    - \frac{\dt^2}{12} \left(c^4\tu_{xxxx} + c^2 f_{xx} + f_{tt} \right) 
    \nonumber \\
    & & + \ O(\dx^4) + O(\dt^4).
\eea
Thus, we can eliminate the leading-order term in the discretization
error by choosing $\dt = \dx/c$ and adding the correction term 
\beq
  \frac{\dt^2}{12} \left(c^2 f_{xx} + f_{tt} \right) 
\eeq
to the right-hand side of (\ref{eq:wave_eqn_KPY}).

@@

\paragraph{High Accuracy Required for First Time Step}
For the KPY discretization of the wave equation, it is critical that a 
high-order accurate method is used to perform the first time step because 
the error introduced during this step affects the global error at all times.
Specifically, for the error of the KPY scheme to be $O(\dx^p)$, the numerical
solution at the first time step must be accurate to $O(\dx^{p+1})$.  In other 
words, the error introduced by the first time step must be at least one order
of accuracy higher than desired for the overall solution.

We can understand this need for higher-order accuracy of the first time step 
by solving the difference equation
\bea
  \frac{\he^{n+1} - 2 \he^{n} + \he^{n-1}}
       {\dt^2} = \dt^2 \lambda \he^{n}
  \label{eq:error_eqn_normal_mode},
\eea
for the normal modes of the error, 
where $\he$ is the coefficient of an arbitrary normal mode of the spatial
operator for the error $e \equiv u - \tu$.  The solution of this equation 
is~\cite{carrier_pearson_book}
\bea
  \he^n = \he^1 \frac{\kappa_+^n - \kappa_-^n}{\kappa_+ - \kappa_-} 
      + \he^0 \kappa_+ \kappa_- 
        \frac{\kappa_+^{n-1} - \kappa_-^{n-1}}{\kappa_+ - \kappa_-},
  \label{eq:error_eqn_normal_mode_soln}
\eea
where $\kappa_\pm$, the roots of the characteristic equation for 
(\ref{eq:error_eqn_normal_mode}), are given by~\cite{kreiss2002}
\beq
  \kappa_\pm = 1 + \frac{1}{2} \lambda \dt^2 
             \pm \dt \sqrt{\lambda + \frac{\lambda^2 \dt^2}{4}}.
\eeq
Notice that the denominator of both terms in 
(\ref{eq:error_eqn_normal_mode_soln}) is $O(\dt)$.  As a result, the 
global error is always at least one temporal order of accuracy less than the 
error in the initial conditions.  Therefore, the initial errors $e^0$ and 
$e^1$ must be $O(\dt^{p+1})$ in order to for the solution to be $O(\dt^p)$. 
For KPY without OTS selection, this observation indicates that the first
time step should be at least third-order accurate, which is exactly the way
that Kreiss \etal chose to compute the first time step in~\cite{kreiss2002}.
Because the OTS selection transforms KPY into a fourth-order accurate scheme, 
we require that the first time step is at least fifth-order accurate.

It is straightforward to construct a fifth-order approximation for the first 
time step by using a fifth-order Taylor series expansion in time:
\bea
  u^1 &=& u^0 + \dt u_t^0 + \frac{\dt^2}{2} u_{tt}^0
      + \frac{\dt^3}{6} u_{ttt}^0 
   + \frac{\dt^4}{24} u_{tttt}^0
  \nonumber \\
  &=& u^0 + \dt u_t^0 + \frac{\dt^2}{2} \left(c^2 u_{xx}^0 + f\right)
  \nonumber \\
  & & + \frac{\dt^3}{6} \left(c^2 \left(u_t\right)_{xx}^0 + f_t\right)
  \nonumber \\
  & & + \ \frac{\dt^4}{24} \left(c^4 u_{xxxx}^0 + c^2 f_{xx} + f_{tt}\right).
  \label{eq:KPY_fifth_order_first_step}
\eea
Note that sufficiently high-order accurate finite difference stencils may be 
used to compute some of the higher order terms in this expression.  However,
when advancing the solution using the optimal time step, special care must be 
exercised if finite differences are used to compute the derivatives in 
(\ref{eq:KPY_fifth_order_first_step}) because higher-order terms may be 
automatically (implicitly) included by lower-order terms in the expansion.

\subsubsection*{OTS Selection for Variable Coefficient Wave Equation}
We now consider the variable coefficient second-order wave equation
\beq
    u_{tt} - c(x)^2 u_{xx} = f,
    \label{eq:var_coef_wave_eqn}
\eeq
on the domain $0 < x < 1$.  Using the change of variables defined in 
(\ref{eq:change_of_variables}) and (\ref{eq:abar}), we can transform 
(\ref{eq:var_coef_wave_eqn}) into the constant-coefficient wave equation
\beq
  u_{tt} = \cbar^2 u_{yy} - \cbar c' u_y + f,
\eeq
where $c' = dc/dx$.
